---
title: "Extend TapeR by implementing sorting and related aspects of forestry in R"
author: "Christian Vonderach"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
# output: pdf_document
vignette: >
  %\VignetteIndexEntry{TapeS}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to get the R-Package
```{r, eval=FALSE}
remotes::install_gitlab(„vochr/tapes“, build_vignettes = TRUE) 
```

## Basic
This R-Package extends, adapts and realizes the methodological taper curve
development package *TapeR* for the context of the German National Forest
Inventory (NFI). For that purpose new taper curve models were developed on 
extensive data from several German research stations.

## Functionality
The package aims at providing functions to evaluate TapeR-base taper curves with
respect to diameter at given height, height of given diameter, estimating
double bark thickness, providing volumes of total taper curve and parts of it,
estimating total aboveground biomass and components of it. The following section
provide examples on the data structure and how the functions can be used.

### S4-classes
The package establishes two S4-classes, which define requirements on the data
applied to the functions. The *tprTrees*-class obviously defines trees, but 
with a focus on sectional measurements. With that, one can provide as many
diameters (and their measurement heights) to precisely follow its taper curve.
But similarly, one can only provide one diameter and its measurement height for
subsequent use of the package functions.
```{r, eval=FALSE}
help("tprTrees-class")
```

For estimating roundwood assortments, there is an additional S4-class 
*parSort* which defines the sorting parameters.
```{r, eval=FALSE}
help("parSort-class")
```


### handled tree species 
The tree species list is taken from the predecessor taper curve model BDAT. 
Indeed, there are only *eight* taper curve models, but the other tree species
are mapped to these eight. For future releases this might chance if further
data leads to the development of additional taper curve models.
```{r}
require(TapeS)
tprSpeciesCode(inSp = NULL, outSp = NULL)
```

### species mapping
Beside the taper curve models, mapping between species is required also for 
double bark thickness, biomass functions (total aboveground and component 
biomass), minimum cutting diameter, the amount of crown wood in deciduous tree
species and the volume for small trees according to FAO.

```{r}
BaMap(Ba = NULL, type = NULL)
```

### Defining a *tprTrees* object
```{r}
require(TapeS)
obj <- tprTrees(spp=c(1, 3),
                Hm=list(c(1.3, 5, 7), c(1.3)),
                Dm=list(c(27, 23.5, 22.4), c(27)),
                Ht=c(27, 30))
Hx <- c(1.3, 5, 7)
```

### Diameter estimation
Diameter estimation from the developed taper curves via the empirical best 
linear unbiased predictor (EBLUP, see Kublin et al. (2013)) not necessarily  
passed through the observations. This behaviour can be forced if the residual
variance is set to zero. The package offers to set an option 'TapeS_R0' which,
by default, is set to FALSE and uses full information from the models.

If one wants to force the taper curves through given (measured) diameters,
one can do so by setting TapeS::setTapeSoptions(). See further down.
```{r}
tprDiameter(obj, Hx = Hx) # R0=TRUE, taper curve through measurements
tprDiameter(obj, Hx = Hx, bark = FALSE)
tprDiameter(obj, Hx = Hx, interval = "prediction")
```

The diameter are evaluated at position Hx for each tree. If only one diameter
is required, e.g. for relative height, the parameter cp=FALSE (no cartesian 
product) can be used
```{r}
tprDiameter(obj, Hx = 0.3*Ht(obj), interval = "prediction", cp=FALSE)
```


### Height estimation
```{r}
tprHeight(obj, Dx = c(10, 9, 8, 7))
tprHeight(obj, Dx = c(10, 9, 8, 7), bark = FALSE)
```

### Estimation of double bark thickness
```{r}
tprBark(obj, Hx = c(1, 2, 3))
```


### Volume estimation
```{r}
tprVolume(obj) # default is Vfm
tprVolume(obj, AB = list(A=0, B=7), iAB=c("h", "dob"), bark=TRUE) # same
Vfm(obj) # wrapper
VolR(obj) # wrapper
Efm(obj, stH = 0.01) # default
VolE(obj)
VolFAO(obj)
Vfm_phys(obj) # takes a while
Efm_phys(obj)
tprVolume(obj, AB = list(A=0.01*Ht(obj), B=7, sl=0.01), iAB = c("H", "Dob"), bark=FALSE)
```

### Estimation of roundwood assortments
```{r}
tprAssortment(obj) ## default assortment parameters
pars <- parSort(stH=0.2, Lxh=c(1, 1.5), fixN=2, fixL=4)
tprAssortment(obj, pars = pars)
```


### Biomass estimation
#### total aboveground biomass
```{r}
tprBiomass(obj) # bwi-biomass
```

#### component biomass
For this package the biomass functions of Vonderach et al. (2018) have been 
extended to include predictors D03 and crown length for the prediction of 
several components. Although these models can estimate absolute biomass, in this
context they are used to predict the share of each component, which is 
multiplied by total aboveground biomass.
```{r}
tprBiomass(obj, component = c("sw", "sb", "ndl")) 
tprBiomass(obj, component = c("all"))
```


## Option: Taper curve evaluation
In TapeR-package, the taper curve of a tree with given diameter- and height 
measurements is as a default realised using the estimated best linear unbiased 
predictor (see Kublin et al. 2013, p.987). This means, that all diameters and 
volume predictions are estimated using the intrinsic error structure of the 
fitted taper curve model. Hence the taper curve not necessarily passes through 
the measured diameter.

For the possibility of forcing the taper curves through the measured diameters,
the TapeR-package has been extended to allow for explicitly determining the 
assumed residual variance. See *?TapeR::resVar*.

To apply a modification of the residual variance in TapeS, one can set a 
special package option (all called functions react to this option) or define the
options as a parameter in each function. The package option is set by
*TapeS_Rfn = list(fn="zero")*, here with a zero-function to force interpolation.
```{r, eval=TRUE}
setTapeSoptions(Rfn = list(fn="zero"))
tprDiameter(obj, Hx=1.3)
setTapeSoptions(Rfn = list(fn="sig2"))
tprDiameter(obj, Hx=1.3)
```
By default, this option is set to "sig2" on load of the package, hence, the
residual variance of the model is used (default behaviour of TapeR). Other 
functions are also available, e.g. "linear", "laglinear", "biliniar", "dnorm" 
and "dlnorm". Most function allow for defining a point of zero variance. The 
latter two are based on the normal and log-normal distribution and allow for a
standard deviation term.

Please note: setting a zero variance should not be overstrained. It is mainly 
implemented to preserve diameter in breast height (dbh) measurements for trees 
of e.g. inventory data. If one is about to apply the TapeS-functions to 
sectional data, where a lot of diameter measurements are available, the 
*TapeS_Rfn*-Option should left as is or set to *sig2* via
TapeS::setTapeSoptions(Rfn = list(fn="sig2")).

## Option: improve monotonicity of taper curves
Depending on the size and measurements of (especially small) trees, it sometimes
might happen, that the calibrated taper curves have an increasing diameter
sequence at the stem foot:
```{r}
plot(tprTrees(spp=3, Dm=7.9, Hm=1.3, Ht=12, ), mono=FALSE)
```

To avoid such implausible taper curves, internally an additional support 
"measurement" at 1% of tree height is added to the set of measurements. This
extra measurement is based on the ratio between 5% and 1% of tree height of the
population average curve. With that, taper curves usually are monotonically 
decreasing:
```{r}
plot(tprTrees(spp=3, Dm=7.9, Hm=1.3, Ht=12, ), mono=TRUE)
```

By default, *mono* is set to TRUE (must not be given) via options("TapeS_mono"),
hence, all non-monotonic curves (check at initialisation), are amended by an
additional support diameter.


## References
Kublin, E., J. Breidenbach and G. Kaendler (2013). "A flexible stem taper and 
volume prediction method based on mixed-effects B-spline regression." European 
Journal of forest research 132(5-6): 983-997.

Vonderach, C., G. Kändler and C. F. Dormann (2018). "Consistent set of additive 
biomass functions for eight tree species in Germany fit by nonlinear seemingly 
unrelated regression." Annals of Forest Science 75(2): 49.

## Contact
christian.vonderach@forst.bwl.de

Forest Research Institute Baden-Württemberg by order of
Thünen Institute of Forest Ecosystems / Forest resources and climate protection

